\vspace*{\baselineskip}

\begin{center}
    \large\bfseries
    For the Liberty of unlicenc'd Printing.
\end{center}

\vspace{\baselineskip}

\lettrine[lines=3]{\bfseries T}{ h}ey who to States and Governours
of the Commonwealth direct their Speech, High Court of Parlament, or wanting
ſuch acceſſe in a private condition, write that which they foreſee may advance
the publick good; I ſuppoſe them as at the beginning of no meane endeavour, not
a little alter'd and mov'd inwardly in their mindes: Some with doubt of what
will be the ſucceſſe, others with fear of what will be the cenſure; ſome with
hope, others with confidence of what they have to ſpeake.  And me perhaps each
of theſe diſpoſitions, as the ſubject was whereon I enter'd, may have at other
times variouſly affected; and likely might in theſe formoſt expreſſions now alſo
diſcloſe which of them ſway'd moſt, but that the very attempt of this addreſſe
thus made, and the thought of whom it hath recourſe to, hath got the power
within me to a paſſion, farre more welcome then incidentall to a Preface.  Which
though I ſtay not to confeſſe ere any aske, I ſhall be blameleſſe, if it be no
other, then the joy and gratulation which it brings to all who wiſh and promote
their Countries liberty; whereof this whole Diſcourſe propos'd will be a
certaine teſtimony, if not a Trophey.  For this is not the liberty which wee can
hope, that no grievance ever ſhould ariſe in the Commonwealth, that let no man
in this World expect; but when complaints are freely heard, deeply conſider'd
and ſpeedily reform'd, then is the utmoſt bound of civill liberty attain'd, that
wiſe men looke for.  To which if I now manifeſt by the very ſound of this which
I ſhall utter, that wee are already in good part arriv'd, and yet from ſuch a
steepe disadvantage of tyranny and superstition grounded into our principles as
was beyond the manhood of a \textit{Roman} recovery, it will bee attributed
firſt, as is moſt due, to the ſtrong aſſiſtance of God our deliverer, next to
your faithfull guidance and undaunted Wiſdome, Lords and Commons of
\textit{England}.  Neither is it in Gods eſteeme the diminution of his glory,
when honourable things are ſpoken of good men and worthy Magistrates; which if I
now firſt ſhould begin to doe, after ſo fair a progreſſe of your laudable deeds,
and ſuch a long obligement upon the whole Realme to your indefatigable virtues,
I might be juſtly reckn'd among the tardieſt, and the unwillingeſt of them that
praiſe yee.  Nevertheleſſe there being three principall things, without which
all praiſing is but Courtſhip and flattery, Firſt, when that only is prais'd
which is ſolidly worth praiſe: next, when greateſt likelihoods are brought that
ſuch things are truly and really in thoſe perſons to whom they are aſcrib'd, the
other, when he who praiſes, by ſhewing that ſuch his actuall perſwaſion is of
whom he writes, can demonſtrate that he flatters not; the former two of theſe I
have heretofore endeavour'd, reſcuing the employment from him who went about to
impaire your merits with a triviall and malignant \textit{Encomium}; the latter
as belonging chiefly to mine owne acquittall, that whom I ſo extoll'd I did not
flatter, hath been reſerv'd opportunely to this occaſion.  For he who freely
magnifies what hath been nobly done, and fears not to declare as freely what
might be done better, gives ye the beſt cov'nant of his fidelity; and that his
loyaleſt affection and his hope waits on your proceedings.  His higheſt praiſing
is not flattery, and his plaineſt advice is a kinde of praiſing; for though I
ſhould affirme and hold by argument, that it would fare better with truth, with
learning, and the Commonwealth, if one of your publiſht Orders which I ſhould
name, were call'd in, yet at the ſame time it could not but much redound to the
luſtre of your milde and equall Government, when as private perſons are hereby
animated to thinke ye better pleas'd with publick advice, then other ſtatiſts
have been delighted heretofore with publicke flattery.  And men will then ſee
what difference there is between the magnanimity of a trienniall Parlament, and
that jealous hautineſſe of Prelates and cabin Counſellours that uſurpt of late,
when as they ſhall obſerve yee in the midd'ſt of your Victories and ſucceſſes
more gently brooking writt'n exceptions againſt a voted Order, then other
Courts, which had produc't nothing worth memory but the weake oſtentation of
wealth, would have endur'd the leaſt ſignifi'd diſlike at any ſudden
Proclamation.  If I ſhould thus farre preſume upon the meek demeanour of your
civill and gentle greatneſſe, Lords and Commons, as what your publiſht Order
hath directly ſaid, that to gainſay, I might defend my ſelfe with eaſe, if any
ſhould accuſe me of being new or inſolent, did they but know how much better I
find ye eſteem it to imitate the old and elegant humanity of Greece, then the
barbarick pride of a \textit{Hunniſh} and \textit{Norwegian} ſtatelines.  And
out of thoſe ages, to whoſe polite wiſdom and letters we ow that we are not yet
\textit{Gothes} and \textit{Jutlanders}, I could name him who from his private
houſe wrote that diſcourſe to the Parlament of \textit{Athens}, that perſwades
them to change the forme of \textit{Democraty} which was then eſtabliſht.  Such
honour was done in thoſe dayes to men who profeſt the ſtudy of wiſdome and
eloquence, not only in their own Country, but in other Lands, that Cities and
Siniories heard them gladly, and with great reſpect, if they had aught in
publick to admoniſh the State.  Thus did \textit{Dion Prusæus}, a ſtranger and a
privat Orator counſell the \textit{Rhodians} against a former Edict: and I
abound with other like examples, which to ſet heer would be ſuperfluous.  But if
from the induſtry of a life wholly dedicated to ſtudious labours, and thoſe
naturall endowments haply not the worſt for two and fifty degrees of northern
latitude, ſo much muſt be derogated, as to count me not equall to any of thoſe
who had this priviledge, I would obtain to be thought not ſo inferior, as your
ſelves are ſuperior to the moſt of them who receiv'd their counſell: and how
farre you excell them, be aſſur'd, Lords and Commons, there can no greater
teſtimony appear, then when your prudent ſpirit acknowledges and obeyes the
voice of reaſon from what quarter ſoever it be heard ſpeaking; and renders ye as
willing to repeal any Act of your own ſetting forth, as any ſet forth by your
Predeceſſors.

If ye be thus reſolv'd, as it were injury to think ye were not; I know not what
ſhould withhold me from preſenting ye with fit inſtance wherein to ſhew both
that love of truth which ye eminently profeſſe, and that uprightneſſe of your
judgement which is not wont to be partiall to your ſelves; by judging over again
that Order which ye have ordain'd \textit{to regulate Printing}.  \textit{That
no Book, pamphlet, or paper ſhall be henceforth Printed, unleſſe the ſame be
firſt approv'd and licenc't by ſuch}, or at leaſt one of ſuch as ſhall be
thereto appointed.  For that part which preſerves juſtly every mans Copy to
himſelfe, or provides for the poor, I touch not, only wiſh they be not made
pretenſes to abuſe and perſecute honeſt and painfull Men, who offend not in
either of theſe particulars.  But that other clauſe of Licencing Books, which we
thought had dy'd with his brother \textit{quadragesimal} and
\textit{matrimonial} when the Prelats expir'd, I ſhall now attend with ſuch a
Homily, as ſhall lay before ye, firſt the inventors of it to bee thoſe whom ye
will be loath to own; next what is to be thought in generall of reading, what
ever ſort the Books be; and that this Order avails nothing to the ſuppreſſing of
ſcandalous, ſeditious, and libellous Books, which were mainly intended to be
ſuppreſt.  Laſt, that it will be primely to the diſcouragement of all learning,
and the ſtop of Truth, not only by diſexerciſing and blunting our abilities in
what we know already, but by hindring and cropping the diſcovery that might bee
yet further made both in religious and civill Wiſdome.

I deny not, but that it is of greateſt concernment in the Church and
Commonwealth, to have a vigilant eye how Bookes demeane themſelves as well as
men; and thereafter to confine, impriſon, and do ſharpeſt juſtice on them as
malefactors: For Books are not abſolutely dead things, but doe contain a
potencie of life in them to be as active as that ſoule was whoſe progeny they
are; nay they do preſerve as in a violl the pureſt efficacie and extraction of
that living intellect that bred them.  I know they are as lively, and as
vigorouſly productive, as thoſe fabulous Dragons teeth; and being ſown up and
down, may chance to ſpring up armed men.  And yet on the other hand, unleſſe
warineſſe be us'd, as good almoſt kill a Man as kill a good Book; who kills a
Man kills a reaſonable creature, Gods Image; but hee who deſtroyes a good Booke,
kills reaſon it ſelfe, kills the Image of God, as it were in the eye.  Many a
man lives a burden to the Earth; but a good Booke is the pretious life-blood of
a maſter ſpirit, imbalm'd and treaſur'd up on purpoſe to a life beyond life.
'Tis true, no age can reſtore a life, whereof perhaps there is no great loſſe;
and revolutions of ages do not oft recover the loſſe of a rejected truth, for
the want of which whole Nations fare the worſe.  We ſhould be wary therefore
what perſecution we raiſe againſt the living labours of publick men, how we
ſpill that ſeaſon'd life of man preſerv'd and ſtor'd up in Books; ſince we ſee a
kinde of homicide may be thus committed, ſometimes a martyrdome, and if it
extend to the whole impreſſion, a kinde of maſſacre, whereof the execution ends
not in the ſlaying of an elementall life, but ſtrikes at that ethereall and fift
eſſence, the breath of reaſon it ſelfe, ſlaies an immortality rather then a
life.  But leſt I ſhould be condemn'd of introducing licence, while I oppoſe
Licencing, I refuſe not the paines to be ſo much Hiſtoricall, as will ſerve to
ſhew what hath been done by ancient and famous Commonwealths, against this
diſorder, till the very time that this project of licencing crept out of the
\textit{Inquisition}, was catcht up by our Prelates, and hath caught ſome of our
Presbyters.

In \textit{Athens} where Books and Wits were ever buſier then in any other part
of \textit{Greece}, I finde but only two ſorts of writings which the Magiſtrate
car'd to take notice of; thoſe either blaſphemous and Atheiſticall, or
Libellous.  Thus the Books of \textit{Protagoras} were by the Iudges of
\textit{Areopagus} commanded to be burnt, and himſelfe baniſht the territory for
a diſcourſe begun with his confeſſing not to know \textit{whether there were
gods, or whether not}: And againſt defaming, it was decreed that none ſhould be
traduc'd by name, as was the manner of \textit{Vetus Comœdia}, whereby we may
gueſſe how they cenſur'd libelling: And this courſe was quick enough, as
\textit{Cicero} writes, to quell both the deſperate wits of other Atheiſts, and
the open way of defaming, as the event ſhew'd.  Of other ſects and opinions,
though tending to voluptuouſneſſe, and the denying of divine providence, they
tooke no heed.  Therefore we do not read that either \textit{Epicurus}, or that
libertine ſchool of \textit{Cyrene}, or what the \textit{Cynick} impudence
utter'd, was ever queſtion'd by the Laws.  Neither is it recorded that the
writings of those old Comedians were ſuppreſſt, though the acting of them were
forbid; and that \textit{Plato} commended the reading of \textit{Aristophanes},
the looſeſt of them all, to his royall ſcholler \textit{Dionyſius}, is commonly
known, and may be excus'd, if holy \textit{Chryſoſtome}, as is reported, nightly
ſtudied ſo much the ſame Author and had the art to cleanſe a ſcurrilous
vehemence into the ſtile of a rouſing Sermon.  That other leading city of
\textit{Greece}, \textit{Lacedæmon}, conſidering that \textit{Lycurgus} their
Law-giver was ſo addicted to elegant learning, as to have been the firſt that
brought out of \textit{Ionia} the ſcatter'd workes of \textit{Homer}, and ſent
the poet \textit{Thales} from \textit{Creet} to prepare and mollifie the
\textit{Spartan} ſurlineſſe with his ſmooth ſongs and odes, the better to plant
among them law and civility, it is to be wonder'd how muſeleſſe and unbookiſh
they were, minding nought but the feats of Warre.  There needed no licencing of
Books among them for they diſlik'd all, but their owne \textit{Laconick
Apothegms}, and took a ſlight occaſion to chaſe \textit{Archilochus} out of
their City, perhaps for compoſing in a higher ſtraine then their own ſouldierly
ballats and roundels could reach to: Or if it were for his broad verſes, they
were not therein ſo cautious, but they were as diſſolute in their promiſcuous
converſing; whence \textit{Euripides} affirmes in \textit{Andromache}, that
their women were all unchaſte.  Thus much may give us light after what ſort
Bookes were prohibited among the Greeks.  The Romans alſo for many ages train'd
up only to a military roughnes, reſembling moſt the \textit{Lacedæmonian} guiſe,
knew of learning little but what their twelve Tables, and the \textit{Pontifick}
College with their \textit{Augurs} and \textit{Flamins} taught them in Religion
and Law, ſo unacquainted with other learning, that when \textit{Carneades} and
\textit{Critolaus}, with the \textit{Stoick Diogenes} comming Embaſſadors to
\textit{Rome}, tooke thereby occaſion to give the City a taſt of their
Philoſophy, they were ſuſpected for ſeducers by no leſſe a man then
\textit{Cato} the Cenſor, who mov'd it in the Senat to diſmiſſe them ſpeedily,
and to baniſh all ſuch \textit{Attick} bablers out of \textit{Italy}.  But
\textit{Scipio} and others of the nobleſt Senators withſtood him and his old
\textit{Sabin} auſterity; honour'd and admir'd the men; and the Cenſor himſelf
at laſt in his old age fell to the ſtudy of that whereof before hee was ſo
ſcrupulous.  And yet at the ſame time \textit{Nævius} and \textit{Plautus} the
firſt Latine comedians had fill'd the City with all the borrow'd Scenes of
\textit{Menander} and \textit{Philemon}.  Then began to be conſider'd there alſo
what was to be don to libellous books and Authors; for \textit{Nævius} was
quickly caſt into priſon for his unbridl'd pen, and releas'd by the
\textit{Tribunes} upon his recantation: We read alſo that libels were burnt, and
the makers puniſht by \textit{Augustus}.  The like ſeverity no doubt was us'd if
ought were impiouſly writt'n againſt their eſteemed gods.  Except in theſe two
points, how the world went in Books, the Magiſtrat kept no reckning.  And
therefore \textit{Lucretius} without impeachment verſifies his Epicuriſm to
\textit{Memmius}, and had the honour to be ſet forth the ſecond time by
\textit{Cicero} ſo great a father of the Commonwealth; although himſelfe
diſputes againſt that opinion in his own writings.  Nor was the Satyricall
ſharpneſſe, or naked plainnes of \textit{Lucilius}, or \textit{Catullus}, or
\textit{Flaccus}, by any order prohibited.  And for matters of State, the ſtory
of \textit{Titus Livius}, though it extoll'd that part which \textit{Pompey}
held, was not therefore ſuppreſſt by \textit{Octavius Cæſar} of the other
Faction.  But that \textit{Naſo} was by him baniſht in his old age, for the
wanton Poems of his youth, was but a meer covert of State over ſome ſecret
cauſe: and beſides, the Books were neither baniſht nor call'd in.  From hence we
ſhall meet with little elſe but tyranny in the Roman Empire, that we may not
marvell, if not ſo often bad, as good Books were ſilenc't.  I ſhall therefore
deem to have bin large anough in producing what among the ancients was
puniſhable to write, ſave only which, all other arguments were free to treat on.

By this time the Emperors were become Chriſtians, whoſe diſcipline in this point
I doe not finde to have bin more ſevere then what was formerly in practice.  The
Books of thoſe whom they took to be grand Hereticks were examin'd, refuted, and
condemn'd in the general Councels; and not till then were prohibited, or burnt
by autority of the Emperor.  As for the writings of Heathen authors, unleſſe
they were plaine invectives againſt Chriſtianity, as thoſe of
\textit{Porphyrius} and \textit{Proclus}, they met with no interdict that can be
cited, till about the year 400. in a \textit{Carthaginian} Councel, wherein
Biſhops themſelves were forbid to read the Books of Gentiles, but Hereſies they
might read: while others long before them on the contrary ſcrupl'd more the
Books of Hereticks, then of Gentiles.  And that the primitive Councels and
Biſhops were wont only to declare what Books were not commendable, paſſing no
furder, but leaving it to each ones conſcience to read or to lay by, till after
the yeare 800. is obſerv'd already by \textit{Padre Paolo} the great unmaſker of
the \textit{Trentine} Councel.  After which time the Popes of \textit{Rome},
engroſſing what they pleas'd of Politicall rule into their owne hands, extended
their dominion over mens eyes, as they had before over their judgements, burning
and prohibiting to be read, what they fanſied not; yet ſparing in their
cenſures, and the Books not many which they ſo dealt with: till \textit{Martin}
the 5. by his Bull not only prohibited, but was the firſt that excommunicated
the reading of hereticall Books; for about that time \textit{Wicklef} and
\textit{Huſſe} growing terrible, were they who firſt drove the Papall Court to a
ſtricter policy of prohibiting.  Which cours \textit{Leo} the 10, and his
ſucceſſors follow'd, untill the Councell of Trent, and the Spaniſh Inquiſition
engendring together brought forth, or perfeted thoſe Catalogues, and expurging
Indexes that rake through the entralls of many an old good Author, with a
violation wors then any could be offer'd to his tomb.  Nor did they ſtay in
matters Hereticall, but any ſubject that was not to their palat, they either
condemn'd in a prohibition, or had it ſtrait into the new Purgatory of an Index.
To fill up the meaſure of encroachment, their laſt invention was to ordain that
no Book, pamphlet, or paper ſhould be Printed (as if S.\textit{Peter} had
bequeath'd them the keys of the Preſſe alſo out of Paradiſe) unleſſe it were
approv'd and licenc't under the hands of 2 or 3 glutton Friers.  For example:

\begin{quote}
    Let the Chancellor \textit{Cini} be pleas'd to ſee if in this preſent work
    be contain'd ought that may withſtand the Printing.
\end{quote}

\hfill \textit{Vincent Rabbatta} Vicar of \textit{Florence}.

\begin{quote}
    \hyphenation{Ca-tho-lick}
    I have ſeen this preſent work, and finde nothing athwart the Catholick faith
    and good manners: in witneſſe whereof I have given, \&c.
\end{quote}

\hfill \textit{Nicolò Cini} Chancellor of \textit{Florence}.

\begin{quote}
    Attending the precedent relation, it is allow'd that this preſent work of
    \textit{Davanzati} may be Printed.
\end{quote}

\hfill \textit{Vincent Rabbatta}, \&c.

\begin{quote}
    It may be printed, \textit{July} 15.
\end{quote}

\hfill Friar \textit{Simon Mompei d'Amelia} Chancellor of the holy office in
\textit{Florence}.

\vspace{\baselineskip}

Sure they have a conceit, if he of the bottomleſſe pit had not long ſince broke
priſon, that this quadruple exorciſm would barre him down.  I fear their next
deſigne will be to get into their cuſtody the licencing of that which they ſay
\textit{Claudius}* \marginpar{*Quo veniam daret flatum \\ crepitumque ventris in
\\ convivio emittendi. \\ Sueton. in Claudio.} intended, but went not through
with.  Voutſafe to ſee another of their forms the Roman ſtamp:

\begin{quote}
    \textit{Imprimatur}, If it ſeem good to the reverend Maſter of the holy
    Palace.
\end{quote}

\hfill \textit{Belcaſtro} Vicegerent.

\begin{quote}
    \textit{Imprimatur}
\end{quote}

\hfill Friar \textit{Nicolò Rodolphi} Maſter of the holy Palace.

\vspace{\baselineskip}

Sometimes 5 \textit{Imprimaturs} are ſeen together dialogue-wiſe in the Piatza
of one Title page, complementing and ducking each to other with their ſhav'n
reverences, whether the Author, who ſtands by in perplexity at the foot of his
Epiſtle, ſhall to the Preſſe or to the ſpunge.  Theſe are the prety
reſponſories, theſe are the dear Antiphonies, that ſo bewitcht of late our
Prelats, and their Chaplaines with the goodly Eccho they made; and beſotted us
to the gay imitation of a lordly Imprimatur, one from Lambeth houſe, another
from the Weſt end of \textit{Pauls}; ſo apiſhly Romanizing, that the word of
command ſtill was ſet downe in Latine; as if the learned Grammaticall pen that
wrote it, would caſt no ink without Latine: or perhaps, as they thought, becauſe
no vulgar tongue was worthy to expreſſe the pure conceit of an
\textit{Imprimatur}; but rather, as I hope, for that our Engliſh, the language
of men ever famous, and formoſt in the atchievements of liberty, will not eaſily
finde ſervile letters anow to ſpell ſuch a dictatorie preſumption in Engliſh.
And thus ye have the Inventors and the originall of Book-licencing ript up and
drawn as lineally as any pedigree.  We have it not, that can be heard of, from
any ancient State, or politie, or Church, nor by any Statute left us by our
Anceſtors elder or later; nor from the moderne cuſtom of any reformed Citty, or
Church abroad; but from the moſt Antichriſtian Councel and the moſt tyrannous
Inquiſition that ever inquir'd.  Till then Books were ever as freely admitted
into the World as any other birth; the iſſue of the brain was no more ſtifl'd
then the iſſue of the womb: no envious \textit{Juno} ſate cros-leg'd over the
nativity of any mans intellectuall off ſpring; but if it prov'd a Monſter, who
denies, but that it was juſtly burnt, or ſunk into the Sea.  But that a Book in
wors condition then a peccant ſoul, ſhould be to ſtand before a Jury ere it be
borne to the World, and undergo yet in darkneſſe the judgement of
\textit{Radamanth} and his Collegues, ere it can paſſe the ferry backward into
light, was never heard before, till that myſterious iniquity, provokt and
troubl'd at the firſt entrance of Reformation, ſought out new limbo's and new
hells wherein they might include our Books alſo within the number of their
damned.  And this was the rare morſell ſo officiouſly ſnatcht up, and ſo
ilfavourdly imitated by our inquiſiturient Biſhops, and the attendant minorites
their Chaplains.  That ye like not now theſe moſt certain Authors of this
licencing order, and that all ſiniſter intention was farre diſtant from your
thoughts, when ye were importun'd the paſſing it, all men who know the integrity
of your actions, and how ye honour Truth, will clear yee readily.

But ſome will ſay, What though the inventors were bad, the thing for all that
may be good?  It may ſo; yet if that thing be no ſuch deep invention, but
obvious, and eaſie for any man to light on, and yet beſt and wiſeſt
Commonwealths through all ages, and occaſions have foreborne to uſe it, and
falſeſt ſeducers, and oppreſſors of men were the firſt who tooke it up, and to
no other purpoſe but to obſtruct and hinder the firſt approach of Reformation; I
am of thoſe who beleeve, it will be a harder alchymy then \textit{Lullius} ever
knew, to ſublimat any good uſe out of ſuch an invention.  Yet this only is what
I requeſt to gain from this reaſon, that it may be held a dangerous and
ſuſpicious fruit, as certainly it deſerves, for the tree that bore it, untill I
can diſſect one by one the properties it has.  But I have firſt to finiſh, as
was propounded, what is to be thought in generall of reading Books, what ever
ſort they be, and whether be more the benefit, or the harm that thence proceeds?

Not to inſiſt upon the examples of \textit{Moſes}, \textit{Daniel} \& Paul, who
were skilfull in all the learning of the Ægyptians, Caldeans, and Greeks, which
could not probably be without reading their Books of all ſorts; in \textit{Paul}
eſpecially, who thought it no defilement to inſert into holy Scripture the
ſentences of three Greek Poets, and one of them a Tragedian, the queſtion was,
notwithſtanding ſometimes controverted among the Primitive Doctors, but with
great odds on that ſide which affirm'd it both lawfull and profitable, as was
then evidently perceiv'd, when \textit{Julian} the Apoſtat, and ſuttleſt enemy
to our faith, made a decree forbidding Chriſtians the ſtudy of heathen learning:
for, ſaid he, they wound us with our own weapons, and with our owne arts and
ſciences they overcome us.  And indeed the Chriſtians were put ſo to their
ſhifts by this crafty means, and ſo much in danger to decline into all
ignorance, that the two \textit{Apollinarii} were fain as a man may ſay, to coin
all the ſeven liberall Sciences out of the Bible, reducing it into divers forms
of Orations, Poems, Dialogues, ev'n to the calculating of a new Chriſtian
grammar.  But, ſaith the Hiſtorian \textit{Socrates}, The providence of God
provided better then the induſtry of \textit{Apollinarius} and his ſon, by
taking away that illiterat law with the life of him who devis'd it.  So great an
injury they then held it to be depriv'd of \textit{Hellenick} learning; and
thought it a perſecution more undermining, and ſecretly decaying the Church,
then the open cruelty of \textit{Decius} or \textit{Dioclesian}.  And perhaps it
was the ſame politick drift that the Divell whipt St. \textit{Jerom} in a lenten
dream, for reading \textit{Cicero}; or elſe it was a fantaſm bred by the feaver
which had then ſeis'd him.  For had an Angel bin his diſcipliner, unleſſe it
were for dwelling too much upon Ciceronianiſms, \& had chaſtiz'd the reading,
not the vanity, it had bin plainly partiall; firſt to correct him for grave
\textit{Cicero}, and not for ſcurrill \textit{Plautus}, whom he confeſſes to
have bin reading not long before; next to correct him only, and let ſo many more
ancient Fathers wax old in thoſe pleaſant and florid ſtudies without the laſh of
ſuch a tutoring apparition; inſomuch that \textit{Baſil} teaches how ſome good
uſe may be made of \textit{Margites}, a ſportfull Poem, not now extant, writ by
\textit{Homer}; and why not then of \textit{Morgante}, an Italian Romanze much
to the ſame purpoſe.  But if it be agreed we ſhall be try'd by viſions, there is
a viſion recorded by \textit{Euſebius} far ancienter then this tale of
\textit{Jerom} to the Nun \textit{Euſtochium}, and beſides has nothing of a
feavor in it.  \textit{Dionyſius Alexandrinuſ} was about the year 240, a perſon
of great name in the Church for piety and learning, who had wont to avail
himſelf much againſt hereticks by being converſant in their Books; untill a
certain Presbyter laid it ſcrupulouſly to his conſcience, how he durſt venture
himſelfe among thoſe defiling volumes.  The worthy man loath to give offence
fell into a new debate with himſelfe what was to be thought; when ſuddenly a
viſion ſent from God, it is his own Epiſtle that ſo averrs it, confirm'd him in
theſe words: Read any books what ever come to thy hands, for thou art ſufficient
both to judge aright, and to examine each matter.  To this revelation he
aſſented the ſooner, as he confeſſes, becauſe it was anſwerable to that of the
Apoſtle to the Theſſalonians, Prove all things, hold faſt that which is good.
And he might have added another remarkable ſaying of the ſame Author; To the
pure, all things are pure, not only meats and drinks, but all kinde of knowledge
whether of good or evill; the knowledge cannot defile, nor conſequently the
books, if the will and conſcience be not defil'd.  For books are as meats and
viands are; ſome of good, ſome of evill ſubſtance; and yet God in that
unapocryphall viſion, ſaid without exception, Riſe \textit{Peter}, kill and eat,
leaving the choice to each mans diſcretion.  Wholeſome meats to a vitiated
ſtomack differ little or nothing from unwholeſome; and beſt books to a naughty
mind are not unappliable to occaſions of evill.  Bad meats will ſcarce breed
good nouriſhment in the healthieſt concoction; but herein the difference is of
bad books, that they to a diſcreet and judicious Reader ſerve in many reſpects
to diſcover, to confute, to forewarn, and to illuſtrate.  Wherof what better
witnes can ye expect I ſhould produce, then one of your own now ſitting in
Parlament, the chief of learned men reputed in this land, Mr. \textit{Selden},
whoſe volume of naturall \& national laws proves, not only by great autorities
brought together, but by exquiſite reaſons and theorems almoſt mathematically
demonſtrative, that all opinions, yea errors, known, read, and collated, are of
main ſervice \& aſſiſtance toward the ſpeedy attainment of what is trueſt.  I
conceive therefore, that when God did enlarge the univerſall diet of mans body,
ſaving ever the rules of temperance, he then alſo, as before, left arbitrary the
dyeting and repaſting of our minds; as wherein every mature man might have to
exerciſe his owne leading capacity.  How great a vertue is temperance, how much
of moment through the whole life of man? yet God committs the managing ſo great
a truſt, without particular Law or preſcription, wholly to the demeanour of
every grown man.  And therefore when he himſelf tabl'd the Jews from heaven,
that Omer which was every mans daily portion of Manna, is computed to have bin
more then might have well ſuffic'd the heartieſt feeder thrice as many meals.
For thoſe actions which enter into a man, rather then iſſue out of him, and
therefore defile not, God uſes not to captivat under a perpetuall childhood of
preſcription, but truſts him with the gift of reaſon to be his own chooſer;
there were but little work left for preaching, if law and compulſion ſhould grow
ſo faſt upon thoſe things which hertofore were govern'd only by exhortation.
\textit{Salomon} informs us that much reading is a wearines to the fleſh; but
neither he, nor other inſpir'd author tells us that ſuch, or ſuch reading is
unlawfull: yet certainly had God thought good to limit us herein, it had bin
much more expedient to have told us what was unlawfull, then what was weariſome.
As for the burning of those Epheſian books by St. \textit{Pauls} converts, tis
reply'd the books were magick, the Syriack ſo renders them.  It was a privat
act, a voluntary act, and leaves us to a voluntary imitation: the men in remorſe
burnt thoſe books which were their own; the Magiſtrat by this example is not
appointed; theſe men practiz'd the books, another might perhaps have read them
in ſome ſort uſefully.  Good and evill we know in the field of this World grow
up together almoſt inſeparably; and the knowledge of good is ſo involv'd and
interwoven with the knowledge of evill, and in ſo many cunning reſemblances
hardly to be diſcern'd, that thoſe confuſed ſeeds which were impos'd on
\textit{Pſyche} as an inceſſant labour to cull out, and ſort aſunder, were not
more intermixt.  It was from out the rinde of one apple taſted, that the
knowledge of good and evill as two twins cleaving together leapt forth into the
World.  And perhaps this is that doom which \textit{Adam} fell into of knowing
good and evill, that is to ſay of knowing good by evill.  As therefore the ſtate
of man now is; what wiſdome can there be to chooſe, what continence to forbeare
without the knowledge of evill?  He that can apprehend and conſider vice with
all her baits and ſeeming pleaſures, and yet abſtain, and yet diſtinguiſh, and
yet prefer that which is truly better, he is the true wayfaring Chriſtian.  I
cannot praiſe a fugitive and cloiſter'd vertue, unexercis'd \& unbreath'd, that
never ſallies out and ſees her adverſary, but ſlinks out of the race, where that
immortall garland is to be run for, not without duſt and heat.  Aſſuredly we
bring not innocence into the world, we bring impurity much rather: that which
purifies us is triall, and triall is by what is contrary.  That vertue therefore
which is but a youngling in the contemplation of evill, and knows not the utmoſt
that vice promiſes to her followers, and rejects it, is but a blank vertue, not
a pure; her whiteneſſe is but an excrementall whiteneſſe; Which was the reaſon
why our ſage and ſerious Poet \textit{Spencer}, whom I dare be known to think a
better teacher then \textit{Scotus} or \textit{Aquinas}, deſcribing true
temperance under the person of \textit{Guion}, brings him in with his palmer
through the cave of Mammon, and the bowr of earthly bliſſe that he might ſee and
know, and yet abſtain.  Since therefore the knowledge and ſurvay of vice is in
this world ſo neceſſary to the conſtituting of human vertue, and the ſcanning of
error to the confirmation of truth, how can we more ſafely, and with leſſe
danger ſcout into the regions of ſin and falſity then by reading all manner of
tractats, and hearing all manner of reaſon?  And this is the benefit which may
be had of books promiſcuouſly read.  But of the harm that may reſult hence three
kinds are uſually reckn'd.  Firſt, is fear'd the infection that may ſpread; but
then all human learning and controverſie in religious points muſt remove out of
the world, yea the Bible it ſelfe; for that oftimes relates blaſphemy not
nicely, it deſcribes the carnall ſenſe of wicked men not unelegantly, it brings
in holieſt men paſſionately murmuring againſt providence through all the
arguments of \textit{Epicurus}: in other great diſputes it anſwers dubiouſly and
darkly to the common reader: And ask a Talmudiſt what ails the modeſty of his
marginal Keri, that \textit{Moſes} and all the Prophets cannot perſwade him to
pronounce the textuall Chetiv.  For theſe cauſes we all know the Bible it ſelfe
put by the Papiſt into the firſt rank of prohibited books.  The ancienteſt
Fathers muſt be next remov'd, as \textit{Clement} of \textit{Alexandria}, and
that \textit{Euſebian} book of Evangelick preparation, tranſmitting our ears
through a hoard of heatheniſh obſcenities to receive the Goſpel.  Who finds not
that \textit{Irenæus}, \textit{Epiphanius}, \textit{Jerom}, and others diſcover
more hereſies then they well confute, and that oft for hereſie which is the
truer opinion.  Nor boots it to ſay for theſe, and all the heathen Writers of
greateſt infection, if it muſt be thought ſo, with whom is bound up the life of
human learning, that they writ in an unknown tongue, ſo long as we are ſure
thoſe languages are known as well to the worſt of men, who are both moſt able,
and moſt diligent to inſtill the poiſon they ſuck, firſt into the Courts of
Princes, acquainting them with the choiceſt delights, and criticiſms of ſin.  As
perhaps did that \textit{Petronius} whom \textit{Nero} call'd his
\textit{Arbiter}, the Maſter of his revels; and that notorious ribald of
\textit{Arezzo}, dreaded, and yet dear to the Italian Courtiers.  I name not him
for poſterities ſake, whom \textit{Harry} the 8. nam'd in merriment his Vicar of
hell.  By which compendious way all the contagion that foreine books can infuſe,
will find a paſſage to the people farre eaſier and ſhorter then an Indian
voyage, though it could be ſail'd either by the North of \textit{Cataio}
Eastward, or of \textit{Canada} Weſtward, while our Spaniſh licencing gags the
Engliſh Preſſe never ſo ſeverely.  But on the other ſide that infection which is
from books of controverſie in Religion, is more doubtfull and dangerous to the
learned, then to the ignorant; and yet thoſe books muſt be permitted untoucht by
the licencer.  It will be hard to inſtance where any ignorant man hath bin ever
ſeduc't by Papiſticall book in Engliſh, unleſſe it were commended and expounded
to him by ſome of that Clergy: and indeed all ſuch tractats whether falſe or
true are as the Propheſie of \textit{Iſaiah} was to the \textit{Eunuch}, not to
be \textit{underſtood without a guide}.  But of our Prieſts and Doctors how many
have bin corrupted by ſtudying the comments of Jeſuits and \textit{Sorboniſts},
and how faſt they could transfuſe that corruption into the people, our
experience is both late and ſad.  It is not forgot, ſince the acute and diſtinct
\textit{Arminius} was perverted meerly by the peruſing of a nameleſſe diſcours
writt'n at \textit{Delf}, which at firſt he took in hand to confute.  Seeing
therefore that thoſe books, \& thoſe in great abundance which are likelieſt to
taint both life and doctrine, cannot be ſuppreſt without the fall of learning,
and of all ability in diſputation, and that theſe books of either ſort are moſt
and ſooneſt catching to the learned, from whom to the common people whatever is
hereticall or diſſolute may quickly be convey'd, and that evill manners are as
perfectly learnt without books a thouſand other ways which cannot be ſtopt, and
evill doctrine not with books can propagate, except a teacher guide, which he
might alſo doe without writing, and ſo beyond prohibiting, I am not able to
unfold, how this cautelous enterpriſe of licencing can be exempted from the
number of vain and impoſſible attempts.  And he who were pleaſantly diſpos'd
could not well avoid to lik'n it to the exploit of that gallant man who thought
to pound up the crows by ſhutting his Parkgate.  Beſides another inconvenience,
if learned men be the firſt receivers out of books \& diſpredders both of vice
and error, how ſhall the licencers themſelves be confided in, unleſſe we can
conferr upon them, or they aſſume to themſelves above all others in the Land,
the grace of infallibility, and uncorruptedneſſe?  And again if it be true, that
a wiſe man like a good refiner can gather gold out of the droſſieſt volume, and
that a fool will be a fool with the beſt book, yea or without book, there is no
reaſon that we ſhould deprive a wiſe man of any advantage to his wiſdome, while
we ſeek to reſtrain from a fool, that which being reſtrain'd will be no
hindrance to his folly.  For if there ſhould be ſo much exactneſſe always us'd
to keep that from him which is unfit for his reading, we ſhould in the judgement
of \textit{Ariſtotle} not only, but of \textit{Salomon}, and of our Saviour, not
voutſafe him good precepts, and by conſequence not willingly admit him to good
books; as being certain that a wiſe man will make better uſe of an idle
pamphlet, then a fool will do of ſacred Scripture. 'Tis next alleg'd we muſt not
expoſe ourſelves to temptations without neceſſity, and next to that, not imploy
our time in vain things.  To both theſe objections one anſwer will ſerve, out of
the grounds already laid, that to all men ſuch books are not temptations, nor
vanities; but uſefull drugs and materialls wherewith to temper and compoſe
effective and ſtrong med'cins, which mans life cannot want.  The reſt, as
children and childiſh men, who have not the art to qualifie and prepare theſe
working mineralls, well may be exhorted to forbear, but hinder'd forcibly they
cannot be by all the licencing that Sainted Inquiſition could ever yet contrive;
which is what I promis'd to deliver next, That this order of licencing conduces
nothing to the end for which it was fram'd; and hath almoſt prevented me by
being clear already while thus much hath bin explaining.  See the ingenuity of
Truth, who when ſhe gets a free and willing hand, opens herſelf faſter then the
pace of method and diſcours can overtake her.  It was the task which I began
with, To ſhew that no Nation, or well inſtituted State, if they valu'd books at
all, did ever uſe this way of licencing; and it might be anſwer'd, that this is
a piece of prudence lately diſcover'd.  To which I return, that as it was a
thing ſlight and obvious to think on, ſo if it had bin difficult to finde out,
there wanted not among them long ſince, who ſuggeſted ſuch a cours; which they
not following, leave us a pattern of their judgement, that it was not the not
knowing, but the not approving, which was the cauſe of their not uſing it.
\textit{Plato}, a man of high autority, indeed, but leaſt of all for his
Commonwealth, in the book of his laws, which no City ever yet receiv'd, fed his
fancie with making many edicts to his ayrie Burgomaſters, which they who
otherwiſe admire him, wiſh had bin rather buried and excus'd in the
\textit{genial} cups of an \textit{Academick} night-ſitting.  By which laws he
ſeems to tolerat no kind of learning, but by unalterable decree, conſiſting moſt
of practicall traditions, to the attainment whereof a Library of ſmaller bulk
then his own dialogues would be abundant.  And there alſo enacts that no Poet
ſhould ſo much as read to any privat man, what he had writt'n, untill the Judges
and Law-keepers had ſeen it, and allow'd it: But that \textit{Plato} meant this
Law peculiarly to that Commonwealth which he had imagin'd, and to no other, is
evident.  Why was he not elſe a Law-giver to himſelf, but a tranſgreſſor, and to
be expell'd by his own Magiſtrats; both for the wanton epigrams and dialogues
which he made, and his perpetuall reading of \textit{Sophron Mimus} and
\textit{Ariſtophanes}, books of groſſeſt infamy, and alſo for commending the
latter of them, though he were the malicious libeller of his chief friends, to
be read by the Tyrant \textit{Dionyſius}, who had little need of ſuch traſh to
ſpend his time on?  But that he knew this licencing of Poems had reference and
dependence to many other proviſo's there ſet down in his fancied republic, which
in this world could have no place: and ſo neither he himſelf, nor any Magiſtrat,
or City ever imitated that cours, which tak'n apart from thoſe other collaterall
injunctions, muſt needs be vain and fruitleſſe.  For if they fell upon one kind
of ſtrictneſſe, unleſſe their care were equall to regulat all other things of
like aptnes to corrupt the mind, that ſingle endeavour they knew would be but a
fond labour; to ſhut and fortifie one gate againſt corruption, and be
neceſſitated to leave others round about wide open.  If we think to regulat
Printing, thereby to rectifie manners, we muſt regulat all recreations and
paſtimes, all that is delightful to man.  No muſick muſt be heard, no ſong be
ſet or ſung, but what is grave and \textit{Dorick}.  There muſt be licencing
dancers, that no geſture, motion, or deportment be taught our youth but what by
their allowance ſhall be thought honeſt; for ſuch \textit{Plato} was provided
of; It will ask more then the work of twenty licencers to examin all the lutes,
the violins, and the ghittarrs in every houſe; they muſt not be ſuffer'd to
prattle as they doe, but muſt be licenc'd what they may ſay.  And who ſhall
ſilence all the airs and madrigalls, that whiſper ſoftnes in chambers?  The
Windows alſo, and the \textit{Balcone's} muſt be thought on, there are ſhrewd
books, with dangerous Frontiſpices ſet to ſale; who ſhall prohibit them, ſhall
twenty licencers?  The villages alſo muſt have their viſitors to enquire what
lectures the bagpipe and the rebbeck reads ev'n to the ballatry, and the gammuth
of every \textit{municipal} fidler, for theſe are the Countrymans
\textit{Arcadia's} and his \textit{Monte Mayors}.  Next, what more Nationall
corruption, for which England hears ill abroad, then houſhold gluttony; who
ſhall be the rectors of our daily rioting?  and what ſhall be done to inhibit
the multitudes that frequent thoſe houſes where drunk'nes is ſold and harbour'd?
Our garments alſo ſhould be referr'd to the licencing of ſome more ſober
work-maſters to ſee them cut into a leſſe wanton garb.  Who ſhall regulat all
the mixt converſation of our youth, male and female together, as is the faſhion
of this Country, who ſhall ſtill appoint what ſhall be diſcours'd, what
preſum'd, and no furder?  Laſtly, who ſhall forbid and ſeparat all idle reſort,
all evill company?  Theſe things will be, and muſt be; but how they ſhall be
leſt hurtfull, how leſt enticing, herein conſiſts the grave and governing wiſdom
of a State.  To ſequeſter out of the world into \textit{Atlantick} and
\textit{Eutopian} polities which never can be drawn into uſe, will not mend our
condition; but to ordain wiſely as in this world of evill, in the midd'ſt
whereof God hath plac't us unavoidably.  Nor is it \textit{Plato'ſ} licencing of
books will doe this, which neceſſarily pulls along with it ſo many other kinds
of licencing, as will make us all both ridiculous and weary, and yet fruſtrat;
but thoſe unwritt'n, or at leaſt unconſtraining laws of vertuous education,
religious and civill nurture, which \textit{Plato} there mentions, as the bonds
and ligaments of the Commonwealth, the pillars and the ſuſtainers of every
writt'n Statute; theſe they be which will bear chief ſway in ſuch matters as
theſe, when all licencing will be eaſily eluded.  Impunity and remiſſenes, for
certain are the bane of a Commonwealth, but here the great art lyes to diſcern
in what the law is to bid reſtraint and puniſhment, and in what things
perſwaſion only is to work.  If every action which is good, or evill in man at
ripe years, were to be under pittance, and preſcription, and compulſion, what
were vertue but a name, what praiſe could be then due to well-doing, what
grammercy to be ſober, juſt or continent? many there be that complain of divin
Providence for ſuffering \textit{Adam} to tranſgreſſe, fooliſh tongues! when God
gave him reaſon, he gave him freedom to chooſe, for reaſon is but chooſing; he
had bin elſe a meer artificiall \textit{Adam}, ſuch an \textit{Adam} as he is in
the motions.  We our ſelves eſteem not of that obedience, or love, or gift,
which is of force: God therefore left him free, ſet before him a provoking
object, ever almoſt in his eyes herein conſiſted his merit, herein the right of
his reward, the praiſe of his abſtinence.  Wherefore did he creat paſſions
within us, pleaſures round about us, but that theſe rightly temper'd are the
very ingredients of vertu?  They are not skilfull conſiderers of human things,
who imagin to remove ſin by removing the matter of ſin; for, beſides that it is
a huge heap increaſing under the very act of diminiſhing though ſome part of it
may for a time be withdrawn from ſome perſons, it cannot from all, in ſuch a
univerſall thing as books are; and when this is done, yet the ſin remains
entire.  Though ye take from a covetous man all his treaſure, he has yet one
jewell left, ye cannot bereave him of his covetouſneſſe.  Baniſh all objects of
luſt, ſhut up all youth into the ſevereſt diſcipline that can be exercis'd in
any hermitage, ye cannot make them chaſte, that came not thither ſo: ſuch great
care and wiſdom is requir'd to the right managing of this point.  Suppoſe we
could expell ſin by this means; look how much we thus expell of ſin, ſo much we
expell of vertue: for the matter of them both is the ſame; remove that, and ye
remove them both alike.  This juſtifies the high providence of God, who though
he command us temperance, juſtice, continence, yet powrs out before us ev'n to a
profuſenes all deſirable things, and gives us minds that can wander beyond all
limit and ſatiety.  Why ſhould we then affect a rigour contrary to the manner of
God and of nature, by abridging or ſcanting thoſe means, which books freely
permitted are, both to the triall of vertue and the exerciſe of truth.  It would
be better done to learn that the law muſt needs be frivolous which goes to
reſtrain things, uncertainly and yet equally working to good, and to evill.  And
were I the chooſer, a dram of well-doing ſhould be preferr'd before many times
as much the forcible hindrance of evill-doing.  For God ſure eſteems the growth
and compleating of one vertuous perſon, more then the reſtraint of ten vitious.
And albeit whatever thing we hear or ſee, ſitting, walking, travelling, or
converſing may be fitly call'd our book, and is of the ſame effect that writings
are, yet grant the thing to be prohibited were only books, it appears that this
order hitherto is far inſufficient to the end which it intends.  Do we not ſee,
not once or oftner, but weekly that continu'd Court-libell againſt the Parlament
and City, Printed, as the wet ſheets can witnes, and diſpers't among us, for all
that licencing can doe.  yet this is the prime ſervice a man would think,
wherein this order ſhould give proof of it ſelf.  If it were executed, you'l
ſay.  But certain, if execution be remiſſe or blindfold now, and in this
particular, what will it be hereafter and in other books?  If then the order
ſhall not be vain and fruſtrat, behold a new labour, Lords and Commons, ye muſt
repeal and proſcribe all ſcandalous and unlicenc't books already printed and
divulg'd; after ye have drawn them up into a liſt, that all may know which are
condemn'd, and which not; and ordain that no forrein books be deliver'd out of
cuſtody, till they have bin read over.  This office will require the whole time
of not a few overſeers, and thoſe no vulgar men.  There be alſo books which are
partly uſefull and excellent, partly culpable and pernicious; this work will ask
as many more officials, to make expurgations, and expunctions, that the
Commonwealth of Learning be not damnify'd.  In fine, when the multitude of books
encreaſe upon their hands, ye muſt be fain to catalogue all thoſe Printers who
are found frequently offending, and forbidd the importation of their whole
ſuſpected \textit{typography}.  In a word, that this your order may be exact,
and not deficient, ye muſt reform it perfectly according to the model of
\textit{Trent} and \textit{Sevil}, which I know ye abhorre to doe.  Yet though
ye ſhould condiſcend to this, which God forbid, the order ſtill would be but
fruitleſſe and defective to that end whereto ye meant it.  If to prevent ſects
and ſchiſms, who is ſo unread or ſo uncatechis'd in ſtory, that hath not heard
of many ſects refuſing books as a hindrance, and preſerving their doctrine
unmixt for many ages, only by unwritt'n traditions.  The Chriſtian faith, for
that was once a ſchiſm, is not unknown to have ſpread all over \textit{Aſia},
ere any Goſpel or Epiſtle was ſeen in writing.  If the amendment of manners be
aym'd at, look into Italy and Spain, whether thoſe places be one ſcruple the
better, the honeſter, the wiſer, the chaſter, ſince all the inquiſitionall rigor
that hath been executed upon books.

Another reaſon, whereby to make it plain that this order will miſſe the end it
ſeeks, conſider by the quality which ought to be in every licencer.  It cannot
be deny'd but that he who is made judge to ſit upon the birth, or death of books
whether they may be wafted into this world, or not, had need to be a man above
the common meaſure, both ſtudious, learned, and judicious; there may be elſe no
mean miſtakes in the cenſure of what is paſſable or not; which is alſo no mean
injury.  If he be of ſuch worth as behoovs him, there cannot be a more tedious
and unpleaſing Journey-work, a greater loſſe of time levied upon his head, then
to be made the perpetuall reader of unchoſen books and pamphlets, oftimes huge
volumes.  There is no book that is acceptable unleſſe at certain ſeaſons; but to
be enjoyn'd the reading of that at all times, and in a hand ſcars legible,
whereof three pages would not down at any time in the faireſt Print, is an
impoſition I cannot beleeve how he that values time, and his own ſtudies, or is
but of a ſenſible noſtrill ſhould be able to endure.  In this one thing I crave
leave of the preſent licencers to be pardon'd for ſo thinking: who doubtleſſe
took this office up, looking on it through their obedience to the Parlament,
whoſe command perhaps made all things ſeem eaſie and unlaborious to them; but
that this ſhort triall hath wearied them out already, their own expreſſions and
excuſes to them who make ſo many journeys to ſollicit their licence, are
teſtimony anough.  Seeing therefore thoſe who now poſſeſſe the imployment, by
all evident ſigns wiſh themſelves well ridd of it, and that no man of worth,
none that is not a plain unthrift of his own hours is ever likely to ſucceed
them, except he mean to put himſelf to the ſalary of a Preſſe-corrector, we may
eaſily foreſee what kind of licencers we are to expect hereafter, either
ignorant, imperious, and remiſſe, or baſely pecuniary.  This is what I had to
ſhew, wherein this order cannot conduce to that end, whereof it bears the
intention.

I laſtly proceed from the no good it can do, to the manifeſt hurt it cauſes, in
being firſt the greateſt diſcouragement and affront, that can be offer'd to
learning and to learned men.  It was the complaint and lamentation of Prelats,
upon every leaſt breath of a motion to remove pluralities, and diſtribute more
equally Church revennu's, that then all learning would be for ever daſht and
diſcourag'd.  But as for that opinion, I never found cauſe to think that the
tenth part of learning ſtood or fell with the Clergy: nor could I ever but hold
it for a ſordid and unworthy ſpeech of any Churchman who had a competency left
him.  If therefore ye be loath to diſhearten utterly and diſcontent, not the
mercenary crew of falſe pretenders to learning, but the free and ingenuous ſort
of ſuch as evidently were born to ſtudy, and love lerning for it ſelf, not for
lucre, or any other end, but the ſervice of God and of truth, and perhaps that
laſting fame and perpetuity of praiſe which God and good men have conſented
ſhall be the reward of thoſe whoſe publiſht labours advance the good of mankind,
then know, that ſo far to diſtruſt the judgement \& the honeſty of one who hath
but a common repute in learning, and never yet offended, as not to count him fit
to print his mind without a tutor and examiner, leſt he ſhould drop a ſciſm, or
ſomething of corruption, is the greateſt diſpleaſure and indignity to a free and
knowing ſpirit that can be put upon him.  What advantage is it to be a man over
it is to be a boy at ſchool, if we have only ſcapt the ferular, to come under
the feſcu of an \textit{Imprimatur}? if ſerious and elaborat writings, as if
they were no more then the theam of a Grammar lad under his Pedagogue muſt not
be utter'd without the curſory eyes of a temporizing and extemporizing licencer.
He who is not truſted with his own actions, his drift not being known to be
evill, and ſtanding to the hazard of law and penalty, has no great argument to
think himſelf reputed in the Commonwealth wherin he was born, for other then a
fool or a foreiner.  When a man writes to the world, he ſummons up all his
reaſon and deliberation to aſſiſt him; he ſearches, meditats, is induſtrious,
and likely conſults and conferrs with his judicious friends; after all which
done he takes himſelf to be inform'd in what he writes, as well as any that writ
before him; if in this the moſt conſummat act of his fidelity and ripeneſſe, no
years, no induſtry, no former proof of his abilities can bring him to that ſtate
of maturity, as not to be ſtill miſtruſted and ſuſpected, unleſſe he carry all
his conſiderat diligence, all his midnight watchings, and expence of
\textit{Palladian} oyl, to the haſty view of an unleaſur'd licencer, perhaps
much his younger, perhaps far his inferiour in judgement, perhaps one who never
knew the labour of book-writing, and if he be not repulſt, or ſlighted, muſt
appear in Print like a punie with his guardian, and his cenſors hand on the back
of his title to be his bayl and ſurety, that he is no idiot, or ſeducer, it
cannot be but a diſhonour and derogation to the author, to the book, to the
priviledge and dignity of Learning.  And what if the author ſhall be one ſo
copious of fancie, as to have many things well worth the adding, come into his
mind after licencing, while the book is yet under the Preſſe, which not ſeldom
happ'ns to the beſt and diligenteſt writers; and that perhaps a dozen times in
one book.  The Printer dares not go beyond his licenc't copy; ſo often then muſt
the author trudge to his leav-giver, that thoſe his new inſertions may be viewd;
and many a jaunt will be made, ere that licencer, for it muſt be the ſame man,
can either be found, or found at leiſure; mean while either the Preſſe muſt
ſtand ſtill, which is no ſmall damage, or the author looſe his accurateſt
thoughts, \& ſend the book forth wors then he had made it, which to a diligent
writer is the greateſt melancholy and vexation that can befall.  And how can a
man teach with autority, which is the life of teaching, how can he be a Doctor
in his book as he ought to be, or elſe had better be ſilent, whenas all he
teaches, all he delivers, is but under the tuition, under the correction of his
patriarchal licencer to blot or alter what preciſely accords not with the
hidebound humor which he calls his judgement.  When every acute reader upon the
firſt ſight of a pedantick licence, will be ready with theſe like words to ding
the book a coits diſtance from him, I hate a pupil teacher, I endure not an
inſtructer that comes to me under the wardſhip of an overſeeing fiſt.  I know
nothing of the licencer, but that I have his own hand here for his arrogance;
who ſhall warrant me his judgement?  The State Sir, replies the Stationer, but
has a quick return, The State ſhall be my governours, but not my criticks; they
may be miſtak'n in the choice of a licencer, as eaſily as this licencer may be
miſtak'n in an author: This is ſome common ſtuffe: and he might adde from Sir
\textit{Francis Bacon}, That \textit{ſuch authoriz'd books are but the language
of the times}.  For though a licencer ſhould happ'n to be judicious more then
ordnary, which will be a great jeopardy of the next ſucceſſion, yet his very
office and his commiſſion enjoyns him to let paſſe nothing but what is vulgarly
receiv'd already.  Nay, which is more lamentable, if the work of any deceaſed
author, though never ſo famous in his life time, and even to this day, come to
their hands for licence to be Printed, or Reprinted, if there be found in his
book one ſentence of a ventrous edge, utter'd in the height of zeal, and who
knows whether it might not be the dictat of a divine Spirit, yet not ſuiting
with every low decrepit humor of their own, though it were \textit{Knox}
himſelf, the Reformer of a Kingdom that ſpake it, they will not pardon him their
daſh: the ſenſe of that great man ſhall to all poſterity be loſt, for the
fearfulneſſe or the preſumptuous raſhneſſe of a perfunctory licencer.  And to
what an author this violence hath bin lately done, and in what book of greateſt
conſequence to be faithfully publiſht, I could now inſtance, but ſhall forbear
till a more convenient ſeaſon.  Yet if theſe things be not reſented ſeriouſly
and timely by them who have the remedy in their power, but that ſuch iron moulds
as theſe ſhall have autority to knaw out the choiceſt periods of exquiſiteſt
books, and to commit ſuch a treacherous fraud againſt the orphan remainders of
worthieſt men after death, the more ſorrow will belong to that haples race of
men, whoſe miſfortune it is to have underſtanding.  Henceforth let no man care
to learn, or care to be more then worldly wiſe; for certainly in higher matters
to be ignorant and ſlothfull, to be a common ſtedfaſt dunce will be the only
pleaſant life, and only in requeſt.

And as it is a particular diſeſteem of every knowing perſon alive, and moſt
injurious to the writt'n labours and monuments of the dead, ſo to me it ſeems an
undervaluing and vilifying of the whole Nation.  I cannot ſet ſo light by all
the invention, the art, the wit, the grave and ſolid judgement which is in
England, as that it can be comprehended in any twenty capacities how good
ſoever, much leſſe that it ſhould not paſſe except their ſuperintendence be over
it, except it be ſifted and ſtrain'd with their ſtrainers, that it ſhould be
uncurrant without their manuall ſtamp.  Truth and underſtanding are not ſuch
wares as to be monopoliz'd and traded in by tickets and ſtatutes, and ſtandards.
We muſt not think to make a ſtaple commodity of all the knowledge in the Land,
to mark and licence it like our broad cloath, and our wooll packs.  What is it
but a ſervitude like that impos'd by the Philiſtims, not to be allow'd the
ſharpning of our own axes and coulters, but we muſt repair from all quarters to
twenty licencing forges.  Had any one writt'n and divulg'd erroneous things \&
ſcandalous to honeſt life, miſuſing and forfeiting the eſteem had of his reaſon
among men, if after conviction this only cenſure were adjudg'd him, that he
ſhould never henceforth write, but what were firſt examin'd by an appointed
officer, whoſe hand ſhould be annext to paſſe his credit for him, that now he
might be ſafely read, it could not be apprehended leſſe then a diſgracefull
puniſhment.  Whence to include the whole Nation, and thoſe that never yet thus
offended, under ſuch a diffident and ſuſpectfull prohibition, may plainly be
underſtood what a diſparagement it is.  So much the more, when dettors and
delinquents may walk abroad without a keeper, but unoffenſive books muſt not
ſtirre forth without a viſible jaylor in thir title.  Nor is it to the common
people leſſe then a reproach; for if we be ſo jealous over them, as that we dare
not truſt them with an Engliſh pamphlet, what doe we but cenſure them for a
giddy, vitious, and ungrounded people; in ſuch a ſick and weak eſtate of faith
and diſcretion, as to be able to take nothing down but through the pipe of a
licencer.  That this is care or love of them, we cannot pretend, whenas in thoſe
Popiſh places where the Laity are moſt hated and diſpis'd the ſame ſtrictnes is
us'd over them.  Wiſdom we cannot call it, becauſe it ſtops but one breach of
licence, nor that neither; whenas thoſe corruptions which it ſeeks to prevent,
break in faſter at other dores which cannot be ſhut.

And in concluſion it reflects to the diſrepute of our Miniſters alſo, of whoſe
labours we ſhould hope better, and of the proficiencie which thir flock reaps by
them, then that after all this light of the Goſpel which is, and is to be, and
all this continuall preaching, they ſhould be ſtill frequented with ſuch an
unprincipl'd, unedify'd, and laick rabble, as that the whiffe of every new
pamphlet ſhould ſtagger them out of their catechiſm, and Chriſtian walking.
This may have much reaſon to diſcourage the Miniſters when ſuch a low conceit is
had of all their exhortations, and the benefiting of their hearers, as that they
are not thought fit to be turn'd looſe to three ſheets of paper without a
licencer, that all the Sermons, all the Lectures preacht, printed, vented in
ſuch numbers, and ſuch volumes, as have now wellnigh made all other books
unſalable, ſhould not be armor anough againſt one ſingle \textit{enchiridion},
without the caſtle of St. \textit{Angelo} of an \textit{Imprimatur}.

And leſt ſom ſhould perſwade ye, Lords and Commons, that theſe arguments of
lerned mens diſcouragement at this your order, are meer flouriſhes, and not
reall, I could recount what I have ſeen and heard in other Countries, where this
kind of inquiſition tyrannizes; when I have ſat among their lerned men, for that
honor I had, and bin counted happy to be born in ſuch a place of
\textit{Philoſophic} freedom, as they ſuppos'd England was, while themſelvs did
nothing but bemoan the ſervil condition into which lerning amongſt them was
brought; that this was it which had dampt the glory of Italian wits; that
nothing had bin there writt'n now theſe many years but flattery and fuſtian.
There it was that I found and viſited the famous \textit{Galileo} grown old, a
priſner to the Inquiſition, for thinking in Aſtronomy otherwiſe then the
Franciſcan and Dominican licencers thought.  And though I knew that England then
was groaning loudeſt under the Prelaticall yoak, nevertheleſſe I took it as a
pledge of future happines, that other Nations were ſo perſwaded of her liberty.
Yet was it beyond my hope that thoſe Worthies were then breathing in her air,
who ſhould be her leaders to ſuch a deliverance, as ſhall never be forgott'n by
any revolution of time that this world hath to finiſh.  When that was once
begun, it was as little in my fear, that what words of complaint I heard among
lerned men of other parts utter'd againſt the Inquiſition, the ſame I ſhould
hear by as lerned men at home utterd in time of Parlament againſt an order of
licencing; and that ſo generally, that when I had diſclos'd my ſelf a companion
of their diſcontent, I might ſay, if without envy, that he whom an honeſt
\textit{quæſtorſhip} had indear'd to the \textit{Sicilians}, was not more by
them importun'd againſt \textit{Verres}, then the favourable opinion which I had
among many who honour ye, and are known and reſpected by ye, loaded me with
entreaties and perſwaſions, that I would not deſpair to lay together that which
juſt reaſon ſhould bring into my mind, toward the removal of an undeſerved
thraldom upon lerning.  That this is not therefore the disburdning of a
particular fancie, but the common grievance of all thoſe who had prepar'd their
minds and ſtudies above the vulgar pitch to advance truth in others, and from
others to entertain it, thus much may ſatiſfie.  And in their name I ſhall for
neither friend nor foe conceal what the generall murmur is; that if it come to
inquiſitioning again, and licencing, and that we are ſo timorous of our ſelvs,
and ſo ſuſpicious of all men, as to fear each book, and the ſhaking of every
leaf, before we know what the contents are, if ſome who but of late were little
better then ſilenc't from preaching, ſhall come now to ſilence us from reading,
except what they pleaſe, it cannot be gueſt what is intended by ſom but a ſecond
tyranny over learning: and will ſoon put it out of controverſie that Biſhops and
Presbyters are the ſame to us both name and thing.  That thoſe evills of Prelaty
which before from five or ſix and twenty Sees were diſtributivly charg'd upon
the whole people, will now light wholly upon learning, is not obſcure to us:
whenas now the Paſtor of a ſmall unlearned Pariſh, on the ſudden ſhall be
exalted Archbiſhop over a large dioces of books, and yet not remove, but keep
his other cure too, a myſticall pluraliſt.  He who but of late cry'd down the
ſole ordination of every novice Batchelor of Art, and deny'd ſole juriſdiction
over the ſimpleſt Pariſhioner, ſhall now at home in his privat chair aſſume both
theſe over worthieſt and excellenteſt books and ableſt authors that write them.
This is not, Yee Covnants and Proteſtations that we have made, this is not to
put down Prelaty, this is but to chop an Epiſcopacy; this is but to tranſlate
the Palace \textit{Metropolitan} from one kind of dominion into another, this is
but an old canonicall ſleight of \textit{commuting} our penance.  To ſtartle
thus betimes at a meer unlicenc't pamphlet will after a while be afraid of every
conventicle, and a while after will make a conventicle of every Chriſtian
meeting.  But I am certain that a State govern'd by the rules of juſtice and
fortitude, or a Church built and founded upon the rock of faith and true
knowledge, cannot be ſo puſillanimous.  While things are yet not conſtituted in
Religion, that freedom of writing ſhould be reſtrain'd by a diſcipline imitated
from the Prelats, and learnt by them from the Inquiſition to ſhut us up all
again into the breſt of a licencer, muſt needs give cauſe of doubt and
diſcouragement to all learned and religious men.  Who cannot but diſcern the
finenes of this politic drift, and who are the contrivers; that while Biſhops
were to be baited down, then all Preſſes might be open; it was the peoples
birthright and priviledge in time of Parlament, it was the breaking forth of
light.  But now the Biſhops abrogated and voided out of the Church, as if our
Reformation ſought no more, but to make room for others into their ſeats under
another name, the Epiſcopall arts begin to bud again, the cruſe of truth muſt
run no more oyle, liberty of Printing muſt be enthrall'd again under a
Prelaticall commiſſion of twenty, the privilege of the people nullify'd, and
which is wors, the freedom of learning muſt groan again, and to her old fetters;
all this the Parlament yet ſitting.  Although their own late arguments and
defences againſt the Prelats might remember them that this obſtructing violence
meets for the moſt part with an event utterly oppoſite to the end which it
drives at: inſtead of ſuppreſſing ſects and ſchiſms, it raiſes them and inveſts
them with a reputation: \textit{The puniſhing of wits enhaunces their autority},
ſaith the Vicount St. \textit{Albans}, \textit{and a forbidd'n writing is
thought to be a certain ſpark of truth that flies up in the faces of them who
ſeeke to tread it out}.  This order therefore may prove a nurſing mother to
ſects, but I ſhall eaſily ſhew how it will be a ſtep-dame to Truth: and firſt by
diſinabling us to the maintenance of what is known already.

Well knows he who uſes to conſider, that our faith and knowledge thrives by
exerciſe, as well as our limbs and complexion.  Truth is compar'd in Scripture
to a ſtreaming fountain; if her waters flow not in a perpetuall progreſſion,
they ſick'n into a muddy pool of conformity and tradition.  A man may be a
heretick in the truth; and if he beleeve things only becauſe his Paſtor ſayes
ſo, or the Aſſembly ſo determins, without knowing other reaſon, though his
belief be true, yet the very truth he holds, becomes his hereſie.  There is not
any burden that ſom would gladlier poſt off to another, then the charge and care
of their Religion.  There be, who knows not that there be of Proteſtants and
profeſſors who live and dye in as arrant an implicit faith, as any lay Papiſt of
Loretto.  A wealthy man addicted to his pleaſure and to his profits, finds
Religion to be a traffick ſo entangl'd, and of ſo many piddling accounts, that
of all myſteries he cannot skill to keep a ſtock going upon that trade.  What
ſhould he doe? fain he would have the name to be religious, fain he would bear
up with his neighbours in that.  What does he therefore, but reſolvs to give
over toyling, and to find himſelf out ſom factor, to whoſe care and credit he
may commit the whole managing of his religious affairs; ſom Divine of note and
eſtimation that muſt be.  To him he adheres, reſigns the whole ware-houſe of his
religion, with all the locks and keyes into his cuſtody; and indeed makes the
very perſon of that man his religion; eſteems his aſſociating with him a
ſufficient evidence and commendatory of his own piety.  So that a man may ſay
his religion is now no more within himſelf, but is becom a dividuall movable,
and goes and comes neer him, according as that good man frequents the houſe.  He
entertains him, gives him gifts, feaſts him, lodges him; his religion comes home
at night, praies, is liberally ſupt, and ſumptuouſly laid to ſleep, riſes, is
ſaluted, and after the malmſey, or ſome well ſpic't bruage, and better
breakfaſted then he whoſe morning appetite would have gladly fed on green figs
between \textit{Betheny} and \textit{Ierusalem}, his Religion walks abroad at
eight, and leavs his kind entertainer in the ſhop trading all day without his
religion.

Another ſort there be who when they hear that all things ſhall be order'd, all
things regulated and ſetl'd, nothing writt'n but what paſſes through the
cuſtom-houſe of certain Publicans that have the tunaging and the poundaging of
all free ſpok'n truth, will ſtrait give themſelvs up into your hands, mak'em \&
cut'em out what religion ye pleaſe; there be delights, there be recreations and
jolly paſtimes that will fetch the day about from ſun to ſun, and rock the
tedious year as in a delightfull dream.  What need they torture their heads with
that which others have tak'n ſo ſtrictly, and ſo unalterably into their own
pourveying.  Theſe are the fruits which a dull eaſe and ceſſation of our
knowledge will bring forth among the people.  How goodly, and how to be wiſht
were ſuch an obedient unanimity as this, what a fine conformity would it ſtarch
us all into? doubtles a ſtanch and ſolid peece of frame-work, as any January
could freeze together.

Nor much better will be the conſequence ev'n among the clergy themſelvs; it is
no new thing never heard of before, for a \textit{parochiall} Miniſter, who has
his reward, and is at his \textit{Hercules} pillars in a warm benefice, to be
eaſily inclinable, if he have nothing elſe that may rouſe up his ſtudies, to
finiſh his circuit in an Engliſh concordance and a \textit{topic folio}, the
gatherings and ſavings of a ſober graduatſhip, a \textit{Harmony} and a
\textit{Catena}, treading the conſtant round of certain common doctrinall heads,
attended with their uſes, motives, marks and means, out of which as out of an
alphabet or ſol fa by forming and transforming, joyning and diſ-joyning
variouſly a little book-craft, and two hours meditation might furniſh him
unſpeakably to the performance of more then a weekly charge of ſermoning: not to
reck'n up the infinit helps of interlinearies, breviaries, \textit{ſynopſes},
and other loitering gear.  But as for the multitude of Sermons ready printed and
pil'd up, on every text that is not difficult, our London trading St.
\textit{Thomas} in his veſtry, and adde to boot St. \textit{Martin}, and St.
\textit{Hugh}, have not within their hallow'd limits more vendible ware of all
ſorts ready made: ſo that penury he never need fear of Pulpit proviſion, having
where ſo plenteouſly to refreſh his magazin.  But if his rear and flanks be not
impal'd, if his back dore be not ſecur'd by the rigid licencer, but that a bold
book may now and then iſſue forth, and give the aſſault to ſome of his old
collections in their trenches, it will concern him then to keep waking, to ſtand
in watch, to ſet good guards and ſentinells about his receiv'd opinions, to walk
the round and counter-round with his fellow inſpectors, fearing leſt any of his
flock be ſeduc't, who alſo then would be better inſtructed, better exercis'd and
diſciplin'd.  And God ſend that the fear of this diligence which muſt then be
us'd, doe not make us affect the lazines of a licencing Church.

For if we be ſure we are in the right, and doe not hold the truth guiltily,
which becomes not, if we our ſelves condemn not our own weak and frivolous
teaching, and the people for an untaught and irreligious gadding rout, what can
be more fair, then when a man judicious, learned, and of a conſcience, for ought
we know, as good as theirs that taught us what we know, ſhall not privily from
houſe to houſe, which is more dangerous, but openly by writing publiſh to the
world what his opinion is, what his reaſons, and wherefore that which is now
thought cannot be ſound.  Chriſt urg'd it as wherewith to juſtifie himſelf, that
he preacht in publick; yet writing is more publick then preaching; and more
eaſie to refutation, if need be, there being ſo many whoſe buſineſſe and
profeſſion meerly it is, to be the champions of Truth; which if they neglect,
what can be imputed but their ſloth, or unability?

Thus much we are hinder'd and diſ-inur'd by this cours of licencing toward the
true knowledge of what we ſeem to know.  For how much it hurts and hinders the
licencers themſelves in the calling of their Miniſtery, more then any ſecular
employment, if they will diſcharge that office as they ought, ſo that of
neceſſity they muſt neglect either the one duty or the other, I inſiſt not,
becauſe it is a particular, but leave it to their own conſcience, how they will
decide it there.

There is yet behind of what I purpos'd to lay open, the incredible loſſe, and
detriment that this plot of licencing puts us to, more then if ſom enemy at ſea
ſhould ſtop up all our hav'ns and ports, and creeks, it hinders and retards the
importation of our richeſt Marchandize, Truth: nay it was firſt eſtabliſht and
put in practice by Antichriſtian malice and myſtery on ſet purpoſe to
extinguiſh, if it were poſſible, the light of Reformation, and to ſettle
falſhood; little differing from that policie wherewith the Turk upholds his
\textit{Alcoran}, by the prohibition of Printing.  'Tis not deny'd, but gladly
confeſt, we are to ſend our thanks and vows to heav'n louder then moſt of
Nations, for that great meaſure of truth which we enjoy, eſpecially in thoſe
main points between us and the Pope, with his appertinences the Prelats: but he
who thinks we are to pitch our tent here, and have attain'd the utmoſt proſpect
of reformation, that the mortall glaſſe wherein we contemplate, can ſhew us,
till we come to \textit{beatific} viſion, that man by this very opinion
declares, that he is yet farre ſhort of Truth.

Truth indeed came once into the world with her divine Maſter, and was a perfect
ſhape moſt glorious to look on: but when he aſcended, and his Apoſtles after Him
were laid aſleep, then ſtrait aroſe a wicked race of deceivers, who as that
ſtory goes of the \textit{Ægyptian Typhon} with his conſpirators, how they dealt
with the good \textit{Oſiris}, took the virgin Truth, hewd her lovely form into
a thouſand peeces, and ſcatter'd them to the four winds.  From that time ever
ſince, the ſad friends of Truth, ſuch as durſt appear, imitating the carefull
ſearch that \textit{Iſis} made for the mangl'd body of \textit{Oſiris}, went up
and down gathering up limb by limb ſtill as they could find them.  We have not
yet found them all, Lords and Commons, nor ever ſhall doe, till her Maſters
ſecond comming; he ſhall bring together every joynt and member, and ſhall mould
them into an immortall feature of lovelines and perfection.  Suffer not theſe
licencing prohibitions to ſtand at every place of opportunity forbidding and
diſturbing them that continue ſeeking, that continue to do our obſequies to the
torn body of our martyr'd Saint.  We boaſt our light; but if we look not wiſely
on the Sun it ſelf, it ſmites us into darknes.  Who can diſcern thoſe planets
that are oft \textit{Combuſt}, and thoſe ſtars of brighteſt magnitude that riſe
and ſet with the Sun, untill the oppoſite motion of their orbs bring them to
ſuch a place in the firmament, where they may be ſeen evning or morning.  The
light which we have gain'd, was giv'n us, not to be ever ſtaring on, but by it
to diſcover onward things more remote from our knowledge.  It is not the
unfrocking of a Prieſt, the unmitring of a Biſhop, and the removing him from off
the \textit{Presbyterian} ſhoulders that will make us a happy Nation, no, if
other things as great in the Church, and in the rule of life both economicall
and politicall be not lookt into and reform'd, we have lookt ſo long upon the
blaze that \textit{Zuinglius} and \textit{Calvin} hath beacon'd up to us, that
we are ſtark blind.  There be who perpetually complain of ſchiſms and ſects, and
make it ſuch a calamity that any man diſſents from their maxims.  'Tis their own
pride and ignorance which cauſes the diſturbing, who neither will hear with
meeknes, nor can convince, yet all muſt be ſuppreſſt which is not found in their
\textit{Syntagma}.  They are the troublers, they are the dividers of unity, who
neglect and permit not others to unite thoſe diſſever'd peeces which are yet
wanting to the body of Truth.  To be ſtill ſearching what we know not, by what
we know, ſtill cloſing up truth to truth as we find it (for all her body is
\textit{homogeneal}, and proportionall), this is the golden rule in
\textit{Theology} as well as in Arithmetick, and makes up the beſt harmony in a
Church; not the forc't and outward union of cold, and neutrall, and inwardly
divided minds.

Lords and Commons of England, conſider what Nation it is whereof ye are, and
whereof ye are the governours: a Nation not ſlow and dull, but of a quick,
ingenious, and piercing ſpirit, acute to invent, ſuttle and ſinewy to diſcours,
not beneath the reach of any point the higheſt that human capacity can ſoar to.
Therefore the ſtudies of learning in her deepeſt Sciences have bin ſo ancient,
and ſo eminent among us, that Writers of good antiquity, and ableſt judgement
have bin perſwaded that ev'n the ſchool of \textit{Pythagoras}, and the
\textit{Perſian} wiſdom took beginning from the old Philoſophy of this Iland.
And that wiſe and civill Roman, \textit{Julius Agricola}, who govern'd once here
for \textit{Cæſar}, preferr'd the naturall wits of Britain, before the labour'd
ſtudies of the French.  Nor is it for nothing that the grave and frugal
\textit{Transylvanian} ſends out yearly from as farre as the mountanous borders
of \textit{Ruſſia}, and beyond the \textit{Hercynian} wildernes, not their
youth, but their ſtay'd men, to learn our language, and our \textit{theologic}
arts.  Yet that which is above all this, the favour and the love of heav'n we
have great argument to think in a peculiar manner propitious and propending
towards us.  Why elſe was this Nation chos'n before any other, that out of her
as out of \textit{Sion} ſhould be proclam'd and ſounded forth the firſt tidings
and trumpet of Reformation to all \textit{Europ}.  And had it not bin the
obſtinat perverſenes of our Prelats againſt the divine and admirable ſpirit of
\textit{Wicklef}, to ſuppreſſe him as a ſchiſmatic and \textit{innovator},
perhaps neither the \textit{Bohemian Huſſe} and \textit{Jerom}, no nor the name
of \textit{Luther}, or of \textit{Calvin} had bin ever known: the glory of
reforming all our neighbours had bin compleatly ours.  But now, as our obdurat
Clergy have with violence demean'd the matter, we are become hitherto the lateſt
and backwardeſt Schollers, of whom God offer'd to have made us the teachers.
Now once again by all concurrence of ſigns, and by the generall inſtinct of holy
and devout men, as they daily and ſolemnly expreſſe their thoughts, God is
decreeing to begin ſome new and great period in his Church, ev'n to the
reforming of Reformation it ſelf: what does he then but reveal Himſelf to his
ſervants, and as his manner is, firſt to his Engliſh-men; I ſay as his manner
is, firſt to us, though we mark not the method of his counſels, and are
unworthy.  Behold now this vaſt City: a City of refuge, the manſion houſe of
liberty, encompaſt and ſurrounded with his protection; the ſhop of warre hath
not there more anvils and hammers waking, to faſhion out the plates and
inſtruments of armed Juſtice in defence of beleaguer'd Truth, then there be pens
and heads there, ſitting by their ſtudious lamps, muſing, ſearching, revolving
new notions and idea's wherewith to preſent, as with their homage and their
fealty the approaching Reformation: others as faſt reading, trying all things,
aſſenting to the force of reaſon and convincement.  What could a man require
more from a Nation ſo pliant and ſo prone to ſeek after knowledge.  What wants
there to ſuch a towardly and pregnant ſoile, but wiſe and faithfull labourers,
to make a knowing people, a Nation of Prophets, of Sages, and of Worthies.  We
reck'n more then five months yet to harveſt; there need not be five weeks, had
we but eyes to lift up, the fields are white already.  Where there is much
deſire to learn, there of neceſſity will be much arguing, much writing, many
opinions; for opinion in good men is but knowledge in the making.  Under theſe
fantaſtic terrors of ſect and ſchiſm, we wrong the earneſt and zealous thirſt
after knowledge and underſtanding which God hath ſtirr'd up in this city.  What
ſome lament of, we rather ſhould rejoyce at, ſhould rather praiſe this pious
forwardnes among men, to reaſſume the ill deputed care of their Religion into
their own hands again.  A little generous prudence, a little forbearance of one
another, and ſom grain of charity might win all theſe diligences to joyn, and
unite into one generall and brotherly ſearch after Truth; could we but forgoe
this Prelaticall tradition of crowding free conſciences and Chriſtian liberties
into canons and precepts of men.  I doubt not, if ſome great and worthy ſtranger
ſhould come among us, wiſe to diſcern the mould and temper of a people, and how
to govern it, obſerving the high hopes and aims, the diligent alacrity of our
extended thoughts and reaſonings in the purſuance of truth and freedom, but that
he would cry out as \textit{Pirrhus} did, admiring the Roman docility and
courage, if ſuch were my \textit{Epirots}, I would not deſpair the greateſt
deſign that could be attempted to make a Church or Kingdom happy.  Yet theſe are
the men cry'd out againſt for ſchiſmaticks and ſectaries; as if, while the
Temple of the Lord was building, ſome cutting, ſome ſquaring the marble, others
hewing the cedars, there ſhould be a ſort of irrationall men who could not
conſider there muſt be many ſchiſms and many diſſections made in the quarry and
in the timber, ere the houſe of God can be built.  And when every ſtone is laid
artfully together, it cannot be united into a continuity, it can but be
contiguous in this world; neither can every peece of the building be of one
form; nay rather the perfection conſiſts in this, that out of many moderat
varieties and brotherly diſſimilitudes that are not vaſtly diſproportionall
ariſes the goodly and the gracefull ſymmetry that commends the whole pile and
ſtructure.  Let us therefore be more conſiderat builders, more wiſe in
ſpirituall architecture, when great reformation is expected.  For now the time
ſeems come, wherein \textit{Moses} the great Prophet may ſit in heav'n rejoycing
to ſee that memorable and glorious wiſh of his fulfill'd, when not only our
ſev'nty Elders, but all the Lords people are become Prophets.  No marvell then
though ſome men, and ſome good men too perhaps, but young in goodneſſe, as
\textit{Joſhua} then was, envy them.  They fret, and out of their own weaknes
are in agony, leſt theſe diviſions and ſubdiviſions will undoe us.  The
adverſarie again applauds, and waits the hour, when they have brancht themſelves
out, ſaith he, ſmall anough into parties and partitions, then will be our time.
Fool! he ſees not the firm root, out of which we all grow, though into branches:
nor will be ware untill he ſee our ſmall divided maniples cutting through at
every angle of his ill united and unweildy brigade.  And that we are to hope
better of all theſe ſuppoſed ſects and ſchiſms, and that we ſhall not need that
ſolicitude honeſt perhaps though over timorous of them that vex in this behalf,
but ſhall laugh in the end, at thoſe malicious applauders of our differences, I
have theſe reaſons to perſwade me.

Firſt, when a City ſhall be as it were beſieg'd and blockt about, her navigable
river infeſted, inrodes and incurſions round, defiance and battell oft rumour'd
to be marching up ev'n to her walls, and ſuburb trenches, that then the people,
or the greater part, more then at other times, wholly tak'n up with the ſtudy of
higheſt and moſt important matters to be reform'd, ſhould be diſputing,
reaſoning, reading, inventing, diſcourſing, ev'n to a rarity, and admiration,
things not before diſcourſt or writt'n of, argues firſt a ſingular good will,
contentedneſſe and confidence in your prudent foreſight, and ſafe government,
Lords and Commons; and from thence derives it ſelf to a gallant bravery and well
grounded contempt of their enemies, as if there were no ſmall number of as great
ſpirits among us, as his was, who when Rome was nigh beſieg'd by
\textit{Hanibal}, being in the City, bought that peece of ground at no cheap
rate, whereon \textit{Hanibal} himſelf encampt his own regiment.  Next it is a
lively and cherfull preſage of our happy ſucceſſe and victory.  For as in a
body, when the blood is freſh, the ſpirits pure and vigorous, not only to vital,
but to rationall faculties, and thoſe in the acuteſt, and the perteſt operations
of wit and ſuttlety, it argues in what good plight and conſtitution the body is,
ſo when the cherfulneſſe of the people is ſo ſprightly up, as that it has not
only wherewith to guard well its own freedom and ſafety, but to ſpare, and to
beſtow upon the ſolideſt and ſublimeſt points of controverſie, and new
invention, it betok'ns us not degenerated, nor drooping to a fatall decay, but
caſting off the old and wrincl'd skin of corruption to outlive theſe pangs and
wax young again, entring the glorious waies of Truth and proſperous vertue
deſtin'd to become great and honourable in theſe latter ages.  Methinks I ſee in
my mind a noble and puiſſant Nation rouſing herſelf like a ſtrong man after
ſleep, and ſhaking her invincible locks: Methinks I ſee her as an Eagle muing
her mighty youth, and kindling her undazl'd eyes at the full midday beam;
purging and unſcaling her long abuſed ſight at the fountain it ſelf of heav'nly
radiance; while the whole noiſe of timorous and flocking birds, with thoſe alſo
that love the twilight, flutter about, amaz'd at what ſhe means, and in their
envious gabble would prognoſticat a year of ſects and ſchiſms.

What would ye doe then, ſhould ye ſuppreſſe all this flowry crop of knowledge
and new light ſprung up and yet ſpringing daily in this City, ſhould ye ſet an
\textit{Oligarchy} of twenty ingroſſers over it, to bring a famin upon our minds
again, when we ſhall know nothing but what is meaſur'd to us by their buſhel?
Beleeve it, Lords and Commons, they who counſell ye to ſuch a ſuppreſſing, doe
as good as bid ye ſuppreſſe your ſelves; and I will ſoon ſhew how.  If it be
deſir'd to know the immediat cauſe of all this free writing and free ſpeaking,
there cannot be aſſign'd a truer then your own mild, and free, and human
government; it is the liberty, Lords and Commons, which your own valorous and
happy counſels have purchaſt us, liberty which is the nurſe of all great wits;
this is that which hath rarify'd and enlighten'd our ſpirits like the influence
of heav'n; this is that which hath enfranchis'd, enlarg'd and lifted up our
apprehenſions degrees above themſelves.  Ye cannot make us now leſſe capable,
leſſe knowing, leſſe eagarly purſuing of the truth, unleſſe ye firſt make your
ſelves, that made us ſo, leſſe the lovers, leſſe the founders of our true
liberty.  We can grow ignorant again, brutiſh, formall, and ſlaviſh, as ye found
us; but you then muſt firſt become that which ye cannot be, oppreſſive,
arbitrary, and tyrannous, as they were from whom ye have free'd us.  That our
hearts are now more capacious, our thoughts more erected to the ſearch and
expectation of greateſt and exacteſt things, is the iſſue of your owne vertu
propagated in us; ye cannot ſuppreſſe that unleſſe ye reinforce an abrogated and
mercileſſe law, that fathers may diſpatch at will their own children.  And who
ſhall then ſtick cloſeſt to ye, and excite others? not he who takes up armes for
cote and conduct, and his four nobles of Danegelt.  Although I diſpraiſe not the
defence of juſt immunities, yet love my peace better, if that were all.  Give me
the liberty to know, to utter, and to argue freely according to conſcience,
above all liberties.

What would be beſt advis'd then, if it be found ſo hurtfull and ſo unequall to
ſuppreſſe opinions for the newnes, or the unſutablenes to a cuſtomary
acceptance, will not be my task to ſay; I only ſhall repeat what I have learnt
from one of your own honourable number, a right noble and pious Lord, who had he
not ſacrific'd his life and fortunes to the Church and Commonwealth, we had not
now miſt and bewayl'd a worthy and undoubted patron of this argument.  Ye know
him I am ſure; yet I for honours ſake, and may it be eternall to him, ſhall name
him, the Lord \textit{Brook}.  He writing of Epiſcopacy, and by the way treating
of ſects and ſchiſms, left Ye his vote, or rather now the laſt words of his
dying charge, which I know will ever be of dear and honour'd regard with Ye, ſo
full of meeknes and breathing charity, that next to his laſt teſtament, who
bequeath'd love and peace to his Diſciples, I cannot call to mind where I have
read or heard words more mild and peacefull.  He there exhorts us to hear with
patience and humility thoſe, however they be miſcall'd, that deſire to live
purely, in ſuch a uſe of Gods Ordinances, as the beſt guidance of their
conſcience gives them, and to tolerat them, though in ſome diſconformity to our
ſelves.  The book itſelf will tell us more at large being publiſht to the world,
and dedicated to the Parlament by him who both for his life and for his death
deſerves, that what advice he left be not laid by without peruſall.

And now the time in ſpeciall is, by priviledge to write and ſpeak what may help
to the furder diſcuſſing of matters in agitation.  The temple of \textit{Janus}
with his two \textit{controverſal} faces might now not unſignificantly be ſet
open.  And though all the windes of doctrin were let looſe to play upon the
earth, ſo Truth be in the field, we do injuriouſly, by licencing and prohibiting
to miſdoubt her ſtrength.  Let her and Falſhood grapple; who ever knew Truth put
to the wors, in a free and open encounter.  Her confuting is the beſt and ſureſt
ſuppreſſing.  He who hears what praying there is for light and clearer knowledge
to be ſent down among us, would think of other matters to be conſtituted beyond
the diſcipline of \textit{Geneva}, fram'd and fabric't already to our hands.
Yet when the new light which we beg for ſhines in upon us, there be who envy,
and oppoſe, if it come not firſt in at their caſements.  What a colluſion is
this, whenas we are exhorted by the wiſe man to uſe diligence, \textit{to ſeek
for wiſdom as for hidd'n treaſures} early and late, that another order ſhall
enjoyn us to know nothing but by ſtatute.  When a man hath bin labouring the
hardeſt labour in the deep mines of knowledge, hath furniſht out his findings in
all their equipage, drawn forth his reaſons as it were a battell raung'd,
ſcatter'd and defeated all objections in his way, calls out his adverſary into
the plain, offers him the advantage of wind and ſun, if he pleaſe; only that he
may try the matter by dint of argument, for his opponents then to ſculk, to lay
ambuſhments, to keep a narrow bridge of licencing where the challenger ſhould
paſſe, though it be valour anough in ſouldierſhip, is but weaknes and cowardice
in the wars of Truth.  For who knows not that Truth is ſtrong next to the
Almighty; ſhe needs no policies, nor ſtratagems, nor licencings to make her
victorious, thoſe are the ſhifts and the defences that error uſes againſt her
power: give her but room, \& do not bind her when ſhe ſleeps, for then ſhe
ſpeaks not true, as the old \textit{Proteus} did, who ſpake oracles only when he
was caught \& bound, but then rather ſhe turns herſelf into all ſhapes, except
her own, and perhaps tunes her voice according to the time, as \textit{Micaiah}
did \textit{before} Ahab, untill ſhe be adjur'd into her own likenes.  Yet is it
not impoſſible that ſhe may have more ſhapes then one.  What elſe is all that
rank of things indifferent, wherein Truth may be on this ſide, or on the other,
without being unlike her ſelf.  What but a vain ſhadow elſe is the abolition of
\textit{thoſe ordinances, that hand writing nayl'd to the croſſe}, what great
purchaſe is this Chriſtian liberty which \textit{Paul} ſo often boaſts of.  His
doctrine is, that he who eats or eats not, regards a day, or regards it not, may
doe either to the Lord.  How many other things might be tolerated in peace, and
left to conſcience, had we but charity, and were it not the chief ſtrong hold of
our hypocriſie to be ever judging one another.  I fear yet this iron yoke of
outward conformity hath left a ſlaviſh print upon our necks; the ghoſt of a
linnen decency yet haunts us.  We ſtumble and are impatient at the leaſt
dividing of one viſible congregation from another, though it be not in
fundamentalls; and through our forwardnes to ſuppreſſe, and our backwardnes to
recover any enthrall'd peece of truth out of the gripe of cuſtom, we care not to
keep truth ſeparated from truth, which is the fierceſt rent and diſunion of all.
We doe not ſee that while we ſtill affect by all means a rigid externall
formality, we may as ſoon fall again into a groſſe conforming ſtupidity, a ſtark
and dead congealment of \textit{wood and hay and ſtubble} forc't and frozen
together, which is more to the ſudden degenerating of a Church then many
\textit{ſubdichotomies} of petty ſchiſms.  Not that I can think well of every
light ſeparation, or that all in a Church is to be expected \textit{gold and
ſilver and pretious ſtones}: it is not poſſible for man to ſever the wheat from
the tares, the good fiſh from the other frie; that muſt be the Angels Miniſtery
at the end of mortall things.  Yet if all cannot be of one mind, as who looks
they ſhould be? this doubtles is more wholſome, more prudent, and more Chriſtian
that many be tolerated, rather then all compell'd.  I mean not tolerated Popery,
and open ſuperſtition, which as it extirpats all religions and civill
ſupremacies, ſo it ſelf ſhould be extirpat, provided firſt that all charitable
and compaſſionat means be us'd to win and regain the weak and the miſled: that
alſo which is impious or evil abſolutely either againſt faith or maners no law
can poſſibly permit, that intends not to unlaw it ſelf: but thoſe neighboring
differences, or rather indifferences, are what I ſpeak of, whether in ſome point
of doctrine or of diſcipline, which though they may be many, yet need not
interrupt \textit{the unity of Spirit}, if we could but find among us the
\textit{bond of peace}.  In the mean while if any one would write, and bring his
helpfull hand to the ſlow-moving Reformation which we labour under, if Truth
have ſpok'n to him before others, or but ſeem'd at leaſt to ſpeak, who hath ſo
bejeſuited us that we ſhould trouble that man with asking licence to doe ſo
worthy a deed? and not conſider this, that if it come to prohibiting, there is
not ought more likely to be prohibited then truth it ſelf; whoſe firſt
appearance to our eyes blear'd and dimm'd with prejudice and cuſtom, is more
unſightly and unplauſible then many errors, ev'n as the perſon is of many a
great man ſlight and contemptible to ſee to.  And what doe they tell us vainly
of new opinions, when this very opinion of theirs, that none muſt be heard, but
whom they like, is the worſt and neweſt opinion of all others; and is the chief
cauſe why ſects and ſchiſms doe ſo much abound, and true knowledge is kept at
diſtance from us; beſides yet a greater danger which is in it.  For when God
ſhakes a Kingdome with ſtrong and healthfull commotions to a generall reforming,
'tis not untrue that many ſectaries and falſe teachers are then buſieſt in
ſeducing; but yet more true it is, that God then raiſes to his own work men of
rare abilities, and more then common induſtry not only to look back and reviſe
what hath bin taught heretofore, but to gain furder and goe on, ſome new
enlighten'd ſteps in the diſcovery of truth.  For ſuch is the order of Gods
enlightning his Church, to diſpenſe and deal out by degrees his beam, ſo as our
earthly eyes may beſt ſuſtain it.  Neither is God appointed and confin'd, where
and out of what place theſe his choſen ſhall be firſt heard to ſpeak; for he
ſees not as man ſees, chooſes not as man chooſes, leſt we ſhould devote our
ſelves again to ſet places, and aſſemblies, and outward callings of men;
planting our faith one while in the old Convocation houſe, and another while in
the Chapell at Weſtminſter; when all the faith and religion that ſhall be there
canoniz'd, is not ſufficient without plain convincement, and the charity of
patient inſtruction to ſupple the leaſt bruiſe of conſcience, to edifie the
meaneſt Chriſtian, who deſires to walk in the Spirit, and not in the letter of
human truſt, for all the number of voices that can be there made; no though
\textit{Harry} the 7. himſelf there, with all his leige tombs about him, ſhould
lend them voices from the dead, to ſwell their number.  And if the men be
erroneous who appear to be the leading ſchiſmaticks, what witholds us but our
ſloth, our ſelf-will, and diſtruſt in the right cauſe, that we doe not give them
gentle meeting and gentle diſmiſſions, that we debate not and examin the matter
throughly with liberall and frequent audience; if not for their ſakes, yet for
our own? ſeeing no man who hath taſted learning, but will confeſſe the many
waies of profiting by thoſe who not contented with ſtale receits are able to
manage, and ſet forth new poſitions to the world.  And were they but as the duſt
and cinders of our feet, ſo long as in that notion they may yet ſerve to poliſh
and brighten the armoury of Truth, ev'n for that reſpect they were not utterly
to be caſt away.  But if they be of thoſe whom God hath fitted for the ſpeciall
uſe of theſe times with eminent and ample gifts, and thoſe perhaps neither among
the Prieſts, nor among the Phariſees, and we in the haſt of a precipitant zeal
ſhall make no diſtinction, but reſolve to ſtop their mouths, becauſe we fear
they come with new and dangerous opinions, as we commonly forejudge them ere we
underſtand them, no leſſe then woe to us, while thinking thus to defend the
Goſpel, we are found the perſecutors.

There have bin not a few ſince the beginning of this Parlament, both of the
Presbytery and others who by their unlicen't books to the contempt of an
\textit{Imprimatur} firſt broke that triple ice clung about our hearts, and
taught the people to ſee day: I hope that none of thoſe were the perſwaders to
renew upon us this bondage which they themſelves have wrought ſo much good by
contemning.  But if neither the check that \textit{Moſes} gave to young
\textit{Joſhua}, nor the countermand which our Saviour gave to young
\textit{John}, who was ſo ready to prohibit thoſe whom he thought unlicenc't, be
not anough to admoniſh our Elders how unacceptable to God their teſty mood of
prohibiting is, if neither their own remembrance what evill hath abounded in the
Church by this lett of licencing, and what good they themſelves have begun by
tranſgreſſing it, be not anough, but that they will perſwade, and execute the
moſt \textit{Dominican} part of the Inquiſition over us, and are already with
one foot in the ſtirrup ſo active at ſuppreſſing, it would be no unequall
diſtribution in the firſt place to ſuppreſſe the ſuppreſſors themſelves; whom
the change of their condition hath puft up, more then their late experience of
harder times hath made wiſe.

And as for regulating the Preſſe, let no man think to have the honour of
adviſing ye better then your ſelves have done in that Order publiſht next before
this, that no book be Printed, unleſſe the Printers and the Authors name, or at
leaſt the Printers be regiſter'd.  Thoſe which otherwiſe come forth, if they be
found miſchievous and libellous, the fire and the executioner will be the
timelieſt and the moſt effectuall remedy, that mans prevention can uſe.  For
this \textit{authentic} Spaniſh policy of licencing books, if I have ſaid aught,
will prove the moſt unlicenc't book it ſelf within a ſhort while; and was the
immediat image of a Star-chamber decree to that purpoſe made in thoſe very times
when that Court did the reſt of thoſe her pious works, for which ſhe is now
fall'n from the Starres with \textit{Lucifer}.  Whereby ye may gueſſe what kinde
of State prudence, what love of the people, what care of Religion, or good
manners there was at the contriving, although with ſingular hypocriſie it
pretended to bind books to their good behaviour.  And how it got the upper hand
of your precedent Order ſo well conſtituted before, if we may beleeve thoſe men
whoſe profeſſion gives them cauſe to enquire moſt, it may be doubted there was
in it the fraud of ſome old \textit{patentees} and \textit{monopolizers} in the
trade of book-ſelling; who under pretence of the poor in their Company not to be
defrauded, and the juſt retaining of each man his ſeverall copy, which God
forbid ſhould be gainſaid, brought divers gloſing colours to the Houſe, which
were indeed but colours, and ſerving to no end except it be to exerciſe a
ſuperiority over their neighbours, men who doe not therefore labour in an honeſt
profeſſion to which learning is indetted, that they ſhould be made other mens
vaſſalls.  Another end is thought was aym'd at by ſome of them in procuring by
petition this Order, that having power in their hands, malignant books might the
eaſier ſcape abroad, as the event ſhews.  But of theſe \textit{Sophiſms} and
\textit{Elenchs} of marchandize I skill not: This I know, that errors in a good
government and in a bad are equally almoſt incident; for what Magiſtrate may not
be miſ-inform'd, and much the ſooner, if liberty of Printing be reduc't into the
power of a few; but to redreſſe willingly and ſpeedily what hath bin err'd, and
in higheſt autority to eſteem a plain advertiſement more then others have done a
ſumptuous bribe, is a vertue (honour'd Lords and Commons) anſwerable to Your
higheſt actions, and whereof none can participat but greateſt and wiſeſt men.

\vspace{0.5\baselineskip}

\begin{center}
    \Large\itshape
    The End.
\end{center}
